---
title: "The Layered Grammar of Graphis"
author: "Brice Green"
institute: "University of Texas at Austin"
date: "2020/02/18 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
  lib_dir: libs
css: ['default', 'metropolis-fonts']
nature:
  highlightStyle: github
highlightLines: true
countIncrementalSlides: false
ratio: 16:10
---

```{r setup, include = F}
knitr::opts_chunk$set(
  fig.height = 5, 
  fig.width = 10
)

```

<style type="text/css">
.big {
  font-size: 40px;
}
.remark-slide-content {
  font-size: 30px;
}
.middle {
  margin-top: 47%;
  font-size: 40px;
  font-weight: bold;
}
</style>

# What is a "grammar" of graphics?
.big[
> The fundamental principles
or rules of an art or science
]
-Oxford English Dictionary

---
# What structures are in the grammar?

- Aesthetics: dimensions of representation
  * X axis
  * Y axis
  * Color
- Geometries: forms of representation
  * Bars
  * Points
- Scales: a mapping from a variable to a representation
  * Color
  * Location
  
---
# Using the grammar

These structures are layered on top of one another!

```{r eval = FALSE, message=F, warning = F}
library(ggplot2)
ggplot(iris, aes(x = Sepal.Length, 
                 y = Petal.Length, 
                 color = Species)) +
  geom_point()
```

---
# Using the grammar

These structures are layered on top of one another!

```{r echo = F, message=F, warning = F}
library(ggplot2)
ggplot(iris, aes(x = Sepal.Length, 
                 y = Petal.Length, 
                 color = Species)) +
  geom_point()
```

---
# Breaking this down

This part creates the basic ggplot, the data argument tells it which variables to use:
```
ggplot(data = iris)
```

---
# Breaking this down

The `aes()` call generates the "aesthetics," the dimensions of representation:

```{r eval = F}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species))
```

Aesthetics:
- X: sepal length
- Y: petal length
- Color: species.

---
# Breaking this down

This still has no geometric representations of the data! What happens if we plot this?

```{r echo = F}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species))
```

---
# Breaking this down

The various `geom_` arguments stacked on top give the ways they want the data represented.

```{r eval = F}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species)) +
  geom_point()
```

The geoms "inherit" the aesthetics of the parent function.

---
# Breaking this down

We can stack multiple of these `geom_` arguments:

```{r eval = F}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species)) +
  geom_point() +
  geom_smooth(method = 'lm')
```

---
# Breaking this down

Question: why are there 3 regression lines?

```{r echo = F}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species)) +
  geom_point() +
  geom_smooth(method = 'lm')
```

---
# Changing scales

Scales control the mapping between `geom_` and `aes()`, so if we want to change the axis scale, we would do this with `scale_` arguments.

```{r eval = FALSE}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  scale_x_log10()
```


---
# Changing scales

```{r echo = FALSE}
ggplot(data = iris, 
       aes(x = Sepal.Length, 
           y = Petal.Length, 
           color = Species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  scale_x_log10()
```


---
# Summary

- Aesthetics set up the graph
- Geometries generate representative forms
- Scales control the mappings between aesthetics and forms

---
# Questions?
