---
title: "Tidy Data"
author: "Brice Green"
institute: "University of Texas at Austin"
date: "2020/01/29 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ['default', 'metropolis-fonts']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:10
---

<style type="text/css">
.big {
    font-size: 40px;
}
.remark-slide-content {
  font-size: 30px;
}
.middle {
  margin-top: 47%;
  font-size: 40px;
  font-weight: bold;
}
.small {
  font-size: 15px;
}
</style>


```{r setup, include = F}

library(knitr)
library(dplyr) # made for this stuff!
opts_chunk$set(autodep = T, message = F)

```

# Today:

- _VERY_ brief programming materials & review 
- A framework for structuring data
- Basic computer science concepts behind it
- Examples and practice

---
class: inverse

.middle[
Programming basics
]

---
# Programming basics
## Assignment
* `<-` and `=` assign variables
* case _really_ matters
* variables hold the values you give them

---
# Programming basics
## Objects

* `c()` creates a vector
* `list()` creates a list
* `data.frame()` creates a list of vectors (a table of sorts)
* `matrix()` creates a matrix
* Anything where `x` is a variable that goes `f(x)` is a function

---
# Classes

Classes indicate what kind of thing a variable is. R guesses this automatically, and sometimes it gets it wrong. The basic classes are:

* character for words
* numeric for numbers
* logical for true/false statements


---

class:inverse

.middle[
Tidy Data
]

---
# Tidy Data
> Happy families are all alike; every
unhappy family is unhappy in its own
way.

Leo Tolstoy

---


# Tidy Data
## Definitions

Dataset: collection of _values_, usually numbers or strings

Variable: values that measure the same underlying attribute across units

Observation: all values measured on the same unit

---
# Tidy Data
## Principles

.big[
1. Every column is a variable
2. Every row is an observation
3. Every type of observational unit is a table
]

---
# Not Tidy

|Date       | Alice| Bob| Charlie|
|:----------|-----:|---:|-------:|
|2019-01-01 |     2|   3|      17|
|2020-01-01 |     5|   2|       4|

---
# Tidied

|Date       |Person      | value|
|:----------|:-----------|-----:|
|2019-01-01 |Alice       |     2|
|2020-01-01 |Alice       |     5|
|2019-01-01 |Bob         |     3|
|2020-01-01 |Bob         |     2|
|2019-01-01 |Charlie     |    17|
|2020-01-01 |Charlie     |     4|

---
# Not Tidy

.small[
```{r echo = F}

library(data.table)
tb <- fread("https://raw.githubusercontent.com/hadley/tidy-data/master/data/billboard.csv")


kable(head(tb[, 1:10], 10), format = "markdown")
```
]

---
# Tidied

.small[
|Date| Artist| Length| Name | Date | Week | Rank
|:---|:------|:-----|:---------------|-----:|----:|-----|-----:|
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-02-26 |   1 |  87 | 
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-03-04 |   2 |  82 | 
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-03-11 |   3 |  72 | 
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-03-18 |   4 |  77 | 
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-03-25 |   5 |  87 | 
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-04-01 |   6 |  94 | 
|2000 | 2 Pac | 4:22 | Baby Don't Cry | 2000-04-08 |   7 |  99 | 
|2000 | 2Ge+her | 3:15 | The Hardest Part ... | 2000-09-02 |   1 |  91 | 
|2000 | 2Ge+her | 3:15 | The Hardest Part ... | 2000-09-09 |   2 |  87 | 
]
---
# Working with Tidy Data

The four _verbs_ of data analysis:

* Filter: subsetting or removing observations
* Transform: adding or modifying variables
* Aggregate: collapsing multiple values into a single value
* Sort: re-ordering observations

---
# Working with Tidy Data

Let's see some code examples, then work through a real one.

Create the data.frame (this is what you saw earlier).

```{r, autodep=T}
tidied_data <- data.frame(
  Date = as.Date(rep(c("2019-01-01", "2020-01-01"), 3)),
  Person = c("Alice","Alice", "Bob", "Bob","Charlie" ,"Charlie"), 
  Value = c(2, 5, 3, 2, 17, 4)
)
```
---
# Filtering


```{r eval = F, autodep=T}
library(dplyr) # made for this stuff!
filter(tidied_data, Date == as.Date("2020-01-01"))
```

```{r eval = T, echo = F, autodep=T}
library(dplyr) # made for this stuff!
kable(filter(tidied_data, Date == as.Date("2020-01-01")), format = "markdown")
```

---
# Transforming

What if we want log values? Now it is just one column instead of 3.


```{r eval = F, autodep = T}
tidied_data$LogValue <- log(tidied_data$Value)
```

---
# Aggregating

Average observation of log values by date:

```{r eval = F}

group_by(tidied_data, Date) %>% 
  summarise(Avg = mean(Value))

```

```{r eval = T, echo = F}

group_by(tidied_data, Date) %>% 
  summarise(Avg = mean(Value)) %>% 
  kable(format = "markdown", digits = 2)

```

---

Let's learn how to tidy data, and do some analysis!

